---
title: "Altering or adding variables to data sets"
output: 
  learnr::tutorial:
    progressive: true 
    allow_skip: true 
runtime: shiny_prerendered
description: |
  Adding or altering variables in a dataset is usually necessary to achieve your end goals.
  In the tidyverse this functions are done through the `mutate()` function, which this tutorial
  covers some basic use of. 
---

```{r setup, include=FALSE}
library(tidyquintro)
library(learnr)
library(gradethis)

knitr::opts_chunk$set(
  echo = FALSE,
  exercise.warn_invisible = FALSE
)

# enable code checking
tutorial_options(
  exercise.checker = grade_learnr,
  exercise.lines = 8,
  exercise.reveal_solution = TRUE
)
```


## Challenge 1
### 1a

> Create a column named `bill_ld_ratio` that is the value of `bill_length_mm` divided by `bill_depth_mm`

```{r mutate-1a, exercise = TRUE}
penguins %>% 
  mutate(_ = _ / _) %>% 
  select(species, island, contains("bill"))
```

```{r mutate-1a-solution}
penguins %>% 
  mutate(bill_ld_ratio = bill_length_mm / bill_depth_mm) %>% 
  select(species, island, contains("bill"))
```

```{r mutate-1a-check}
grade_code(
  correct = random_praise(),
  incorrect = random_encouragement()
)
```

```{r mutate-1a-hint-1}
This exercise expects piped data into the mutate function
```

```{r mutate-1a-hint-2}
Make sure you have given the new column the correct name
```

### 1b
> transform the body mass column to the logarithmic scale using the `log` function

```{r mutate-1b, exercise = TRUE}
penguins %>% 
  mutate(body_mass_g_log = __(body_mass_g))
```

```{r mutate-1b-solution}
penguins %>% 
  mutate(body_mass_g_log = log(body_mass_g))
```

```{r mutate-1b-check}
grade_code(
  correct = random_praise(),
  incorrect = random_encouragement()
)
```

```{r mutate-1b-hint-1}
use the `log()` function around body mass.
```

### 1c
> Divide all values in the flipper lenght column with 10, and store it in a variable called flipper_length_cm.

```{r mutate-1c, exercise = TRUE}
penguins %>% 
  mutate(_  = _/10)
```

```{r mutate-1c-solution}
penguins %>% 
  mutate(flipper_length_cm  = flipper_length_mm/10)
```

```{r mutate-1c-check}
grade_code(
  correct = random_praise(),
  incorrect = random_encouragement()
)
```

```{r mutate-1c-hint-1}
use the `log()` function around body mass.
```

## Challenge 2
### 2a
> Adapt the code below to evaluate if body mass is **below** 4.5kg, and assign rows that are TRUE to be "normal" and rows that are FALSE to "large"

```{r mutate-2a, exercise = TRUE}
penguins %>% 
  mutate(body_type = ifelse(body_mass_g _ 4500, "_", "_")) %>% 
  select(species, island, contains("body"))
```

```{r mutate-2a-solution}
penguins %>% 
  mutate(body_type = ifelse(body_mass_g < 4500, "normal", "large")) %>% 
  select(species, island, contains("body"))
```

```{r mutate-2a-check}
grade_code(
  correct = random_praise(),
  incorrect = random_encouragement()
)
```

```{r mutate-2a-hint-1}
After the logical expressions in ifelse, the first value should be the `TRUE` value, and the second the `FALSE`
```

### 2b
```{r mutate-2b, exercise = TRUE}
penguins %>% 
  mutate(biscoe = ifelse(island __ "Biscoe", __, __)) %>% 
  select(species, island, biscoe)
```

```{r mutate-2b-solution}
penguins %>% 
  mutate(biscoe = ifelse(island == "Biscoe", TRUE, FALSE)) %>% 
  select(species, island, biscoe)
```

```{r mutate-2b-check}
grade_code(
  correct = random_praise(),
  incorrect = random_encouragement()
)
```

```{r mutate-2b-hint-2}
Have you used the correct sign for equals `==`
```

```{r mutate-2b-hint-1}
in ifelse, after the logical expressions the first value should be the `TRUE` value, and the second the `FALSE`. 
```

## Challenge 3
### 3a

> Adapt the below code so that penguins with body mass below 3 kg are "petite"

```{r mutate-3a, exercise = TRUE}
penguins %>% 
  mutate(
    body_type = case_when(
      body_mass_g _ 4500 ~ "large",
      body_mass_g _ 3000 ~ "petite",
      _ ~ "normal") # the rest
  ) %>% 
  select(species, island, contains("body"))
```

```{r mutate-3a-solution}
penguins %>% 
  mutate(
    body_type = case_when(
      body_mass_g > 4500 ~ "large",
      body_mass_g < 3000 ~ "petite",
      !is.na(body_mass_g) ~ "normal") # the rest
  ) %>% 
  select(species, island, contains("body"))
```

```{r mutate-3a-check}
grade_code(
  correct = random_praise(),
  incorrect = random_encouragement()
)
```

```{r mutate-3a-hint-1}
Have you used the correct sign for 'larger than'?
```

```{r mutate-3a-hint-2}
Have you used the correct sign for 'smaller than'?
```

```{r mutate-3a-hint-3}
evaluate if something is NA with the `is.na` function. 
```


```{r mutate-3a-hint-4}
To flip a logical, so that TRUE becomes FALSE and vice verse, add `!` in the expression. it means "not".
```


## Challenge 4

> **4a**: Transform all the colmns with milimetres measurements so they are scaled, and the _prefix_ "sc_" to the columns names.


```{r mutate-4a, exercise = TRUE}
penguins %>% 
  mutate(across(ends_with("mm"),
                scale, 
                .names = "{.col}_sc")) %>% 
  select(contains("mm"))
```

```{r mutate-4a-solution}
penguins %>% 
  mutate(across(ends_with("mm"),
                scale, 
                .names = "sc_{.col}")) %>% 
  select(contains("mm"))
```

```{r mutate-4a-check}
grade_code(
  correct = random_praise(),
  incorrect = random_encouragement()
)
```

```{r mutate-4a-hint-1}
Move the change "_sc" to "sc_" and move it before "{.col}".
```

### 4b
> Do the same, but now only for the bill measurements.

```{r mutate-4b, exercise = TRUE}
penguins %>% 
  mutate(across(__("__"),
                scale, 
                .names = "{.col}")) %>% 
  select(__("__"))
```

```{r mutate-4b-solution}
penguins %>% 
  mutate(across(starts_with("mm"),
                scale, 
                .names = "sc_{.col}")) %>% 
  select(starts_with("mm"))
```

```{r mutate-4b-check}
grade_code(
  correct = random_praise(),
  incorrect = random_encouragement()
)
```

```{r mutate-4b-hint-1}
Change ends_with to starts_with
```

### 4c
> Do the same, but now for all the numeric columns.

```{r mutate-4c, exercise = TRUE}
penguins %>% 
  mutate(across(__(is.__),
                scale, 
                .names = "sc_{.col}")) %>% 
  select(__(is.__))
```

```{r mutate-4c-solution}
penguins %>% 
  mutate(across(where(is.numeric),
                scale, 
                .names = "sc_{.col}")) %>% 
  select(where(is.numeric))
```

```{r mutate-4c-check}
grade_code(
  correct = random_praise(),
  incorrect = random_encouragement()
)
```

```{r mutate-4c-hint-1}
Try remembering the `where` function and how it works
```

```{r mutate-4c-hint-2}
try `where(is.numeric)`
```

## Challenge 5

### 5a
> Adapt the code below, so that you de-mean the bill length column. Do it in two steps, first by making a column for the species means, then using that data to de-mean the bill length.

```{r mutate-5a, exercise = TRUE}
penguins %>% 
  group_by(_) %>% 
  mutate(
    bill_length_sp_mean = __(bill_length_mm, na.rm = TRUE),
    bill_length_cent = __ - __
  ) %>% 
  select(species, island, starts_with("bill"))
```

```{r mutate-5a-solution}
penguins %>% 
  group_by(species) %>% 
  mutate(
    bill_length_sp_mean = mean(bill_length_mm, na.rm = TRUE),
    bill_length_cent = bill_length_mm - bill_length_sp_mean
  ) %>% 
  select(species, island, starts_with("bill"))
```

```{r mutate-5a-check}
grade_code(
  correct = random_praise(),
  incorrect = random_encouragement()
)
```

```{r mutate-5a-hint-1}
Did you make sure the column names are correct?
```

### 5b
> Do the same, but now in a single step (i.e. do not store the species mean in its own column)

```{r mutate-5b, exercise = TRUE}
penguins %>% 
  group_by(_) %>% 
  mutate(
    bill_length_cent = 
  ) %>% 
  select(species, island, starts_with("bill"))
```

```{r mutate-5b-solution}
penguins %>% 
  group_by(species) %>% 
  mutate(
    bill_length_cent = bill_length_mm - mean(bill_length_mm, na.rm = TRUE)
  ) %>% 
  select(species, island, starts_with("bill"))
```

```{r mutate-5b-check}
grade_code(
  correct = random_praise(),
  incorrect = random_encouragement()
)
```

```{r mutate-5b-hint-1}
Did you make sure the column names are correct?
```


### 5c
> Based on the code in the previous example, adapt it to be grouped by island in stead of species.

```{r mutate5, exercise = TRUE}
penguins %>% 
  group_by(island) %>% 
  mutate(
    bill_length_sp_max = 
    bill_length_pc = bill_length_mm/max(bill_length_mm, na.rm = TRUE
  ) %>% 
  select(species, island, contains("bill"))
```

```{r mutate5-solution}
penguins %>% 
  group_by(island) %>% 
  mutate(
    bill_length_sp_max = max(bill_length_mm, na.rm = TRUE),
    bill_length_pc = (bill_length_mm/bill_length_sp_max)*100
  ) %>% 
  select(species, island, contains("bill"))
```

```{r mutate5-check}
grade_code(
  correct = random_praise(),
  incorrect = random_encouragement()
)
```

```{r mutate5-hint-1}
Did you make sure the column names are correct?
```

### 5d
>  De-mean all the numerical columns, and give them a suffix of "dm".

```{r mutate-5d, exercise = TRUE}
penguins %>% 
  group_by(__) %>% 
  mutate(
    across(__(__),
           ~ _ - mean(_, na.rm = TRUE),
           .names = "_")
  ) %>% 
  select(species, where(is.numeric))
```

```{r mutate-5d-solution}
penguins %>% 
  group_by(species) %>% 
  mutate(
    across(where(is.numeric),
           ~ .x - mean(.x, na.rm = TRUE),
           .names = "{.col}_dm")
  ) %>% 
  select(species, where(is.numeric))
```

```{r mutate-5d-check}
grade_code(
  correct = random_praise(),
  incorrect = random_encouragement()
)
```

```{r mutate-5d-hint-1}
Do you remember the `where` and `is.numeric` functions?
```

```{r mutate-5d-hint-2}
The across internal placeholder for column values is `.x`
```

```{r mutate-5d-hint-3}
The across internal placeholder for column names is `.col`
```

## Challenge 6
### 6a
>  What is the difference in the minimum body mass for the penguins data sat if the data is grouped by species, and when it is ungrouped?

```{r mutate-6a, exercise = TRUE}
## 6a
penguins %>% 
  group_by(species) %>% 
  
  mutate(
    body_mass_min = min(body_mass_g, na.rm = TRUE)
  )
```

```{r mutate-6a-solution}
## 6a
penguins %>% 
  group_by(species) %>% 
  ungroup() %>% 
  mutate(
    body_mass_min = min(body_mass_g, na.rm = TRUE)
  )
```

```{r mutate-6a-check}
grade_code(
  correct = random_praise(),
  incorrect = random_encouragement()
)
```

```{r mutate-6a-hint-1}
use `ungroup()` after the `group_by` to ungroup the data.
```

We will see more examples of how grouping and then ungrouping can give some real power when working with data.

<!-- ## Quiz -->

<!-- ```{r quiz} -->
<!-- quiz( -->
<!--   question("What functions can you use to subset a data set by rows?", -->
<!--     answer("dplyr's `filter()`", correct = TRUE), -->
<!--     answer("dplyr's `select()`"), -->
<!--     answer("`subset()`", correct = TRUE),
    allow_retry = TRUE -->
<!--   ), -->
<!--   question("What functions can you use to subset a data set by columns", -->
<!--     answer("dplyr's `filter()`"), -->
<!--     answer("dplyr's `select()`", correct = TRUE), -->
<!--     answer("`subset()`", correct = TRUE),
    allow_retry = TRUE -->
<!--   ), -->
<!--   question("If you want to select all columns in data 'df' that contains the string 'something', you can do that by", -->
<!--     answer("`df[grepl('something', names(df))]`", correct = TRUE), -->
<!--     answer("`select(df, starts_with('something')`"), -->
<!--     answer("`df[,'something']`"), -->
<!--     answer("`select(df, contains('something')`", correct = TRUE),
    allow_retry = TRUE -->
<!--   ), -->
<!--     question("If you want to subset rows so that you only have those below 18 years of age, you can do that by", -->
<!--     answer("`df$age < 18`"), -->
<!--     answer("`filter(df, age < 18)`", correct = TRUE), -->
<!--     answer("`df[df$age < 18,]`", correct = TRUE), -->
<!--     answer("`filter(df, age <= 18)`"),
    allow_retry = TRUE -->
<!--   )
<!-- ) -->
<!-- ``` -->



