---
title: "Pivoting data - long and wide formats"
output: 
  learnr::tutorial:
    progressive: true 
    allow_skip: true 
runtime: shiny_prerendered
description: |
  Changing the shape of your data is often necessary for certain other functions
  to work best, or just to tidy up the data. This tutorial covers the two
  pivot functions in the tidyverse to help you get the data from long to wide, and vice versa.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(tidyverse)

knitr::opts_chunk$set(
  echo = FALSE,
  exercise.warn_invisible = FALSE
)

# enable code checking
tutorial_options(
  exercise.checker = grade_learnr,
  exercise.lines = 8,
  exercise.reveal_solution = TRUE
)

penguins_long <- palmerpenguins::penguins %>% 
  mutate(smaple = row_number()) %>% 
  pivot_longer(contains("_"),
               names_to = c("part", "measure" , "unit"),
               names_sep = "_")

penguins_long_simple <- palmerpenguins::penguins %>% 
  mutate(smaple = row_number()) %>% 
  pivot_longer(contains("_"))
```



## Challenge 1
### 1a
> Read in the penguins data set from file.

```{r read-1a, exercise = TRUE}
penguin_path <- palmerpenguins::path_to_file("penguins.csv" )
read.csv(__, header = TRUE)
```

```{r read-1a-solution}
penguin_path <- palmerpenguins::path_to_file("penguins.csv" )
read.csv(penguin_path, header = TRUE)
```

```{r read-1a-check}
grade_code(
  correct = random_praise(),
  incorrect = random_encouragement()
)
```

```{r read-1a-hint-1}
Make sure you provide the object with the path saved in it to read.csv.
```


### 1b
> Read in the penguins raw data set from file

```{r read-1b, exercise = TRUE}
penguin_path <- palmerpenguins::path_to_file(__)
read.csv(__, header = TRUE)
```

```{r read-1b-solution}
penguin_path <- palmerpenguins::path_to_file("penguins_raw.csv" )
read.csv(penguin_path, header = TRUE)
```

```{r read-1b-check}
grade_code(
  correct = random_praise(),
  incorrect = random_encouragement()
)
```

```{r read-1b-hint-1}
Make sure you provide the object with the path saved in it to read.csv.
```

```{r read-1b-hint-2}
Did you change the file to look for to "penguins_raw.csv"?
```

### 1c
> Read in the penguins data set from file and assign it to the object `penguins`

```{r read-1c, exercise = TRUE}
penguin_path <- palmerpenguins::path_to_file(__)
read.csv(__, header = TRUE)
```

```{r read-1c-solution}
penguin_path <- palmerpenguins::path_to_file("penguins.csv" )
penguins <- read.csv(penguin_path, header = TRUE)
```

```{r read-1c-check}
grade_code(
  correct = random_praise(),
  incorrect = random_encouragement()
)
```

```{r read-1c-hint-1}
Make sure you provide the object with the path saved in it to read.csv.
```

```{r read-1c-hint-2}
Did you change the file to look for to "penguins.csv"?
```

```{r read-1c-hint-3}
Make sure you assign the output of `read.csv` to `penguins` using `<-`
```


## Challenge 2
### 2a
> Start by pivoting the penguins data so that all the bill measurements (starts with "bill") are in the same column.

```{r pivot-l-2a, exercise = TRUE}
penguins %>% 
  pivot_longer(_)
```

```{r pivot-l-2a-solution}
penguins %>% 
  pivot_longer(starts_with("bill"))
```

```{r pivot-l-2a-check}
grade_code(
  correct = random_praise(),
  incorrect = random_encouragement()
)
```

```{r pivot-l-2a-hint-1}
Have you selected the columns using `starts_with()`?
```

```{r pivot-l-2a-hint-2}
Do not worry about column naming, just get the measures into a column, and the measure names into another. 
```

### 2b
> Pivot longer all columns with an underscore.

```{r pivot-l-2b, exercise = TRUE}
penguins %>% 
  pivot_longer(_)
```

```{r pivot-l-2b-solution}
penguins %>% 
  pivot_longer(contains("_"))
```

```{r pivot-l-2b-check}
grade_code(
  correct = random_praise(),
  incorrect = random_encouragement()
)
```

```{r pivot-l-2b-hint-1}
Have you selected the columns using `contains()`?
```

```{r pivot-l-2b-hint-2}
Do not worry about column naming, just get the measures into a column, and the measure names into another. 
```

### 2c
> As mentioned, pivot_longer accepts tidy-selectors. Pivot longer all numerical columns.

```{r pivot-l-2c, exercise = TRUE}
penguins %>% 
  pivot_longer(_)
```

```{r pivot-l-2c-solution}
penguins %>% 
  pivot_longer(where(is.numeric))
```

```{r pivot-l-2c-check}
grade_code(
  correct = random_praise(),
  incorrect = random_encouragement()
)
```

```{r pivot-l-2c-hint-1}
Do you remember the `where` and `is.numeric` functions?
```

```{r pivot-l-2c-hint-2}
Try using `where(is.numeric)`
```

## Challenge 3
### 3a
> Pivot longer all the bill measurements, and alter the names in one go, so that there are three columns named "part", "measure" and "unit" after the pivot.

```{r pivot-l-3a, exercise = TRUE}
penguins %>% 
  pivot_longer(_,
               names_to = c(_, _ , _),
               names_sep = _)
```

```{r pivot-l-3a-solution}
penguins %>% 
  pivot_longer(starts_with("bill"),
               names_to = c("part", "measure" , "unit"),
               names_sep = "_")
```

```{r pivot-l-3a-check}
grade_code(
  correct = random_praise(),
  incorrect = random_encouragement()
)
```

```{r pivot-l-3a-hint-1}
Have you selected the columns using `starts_with()`?
```

### 3b
> Pivot longer all the bill measurements, and use the `names_prefix` argument. Give it the string "bill_". What did that do?

```{r pivot-l-3b, exercise = TRUE}
penguins %>% 
  pivot_longer(_,
               names__ = "_")
```

```{r pivot-l-3b-solution}
penguins %>% 
  pivot_longer(starts_with("bill"),
               names_prefix = "bill_")
```

```{r pivot-l-3b-check}
grade_code(
  correct = random_praise(),
  incorrect = random_encouragement()
)
```

```{r pivot-l-3b-hint-1}
be sure to use `names_prefix = "bill_"`
```

### 3c
> Pivot longer all the bill measurements, and use the `names_prefix`, `names_to` and `names_sep` arguments. What do you need to change in `names_to` from the previous example to make it work now that we also use `names_prefix`?

```{r pivot-l-3c, exercise = TRUE}
penguins %>% 
  pivot_longer(starts_with("bill"),
               names_prefix = _,
               names_to = ,
               names_sep = _)
```

```{r pivot-l-3c-solution}
penguins %>% 
  pivot_longer(starts_with("bill"),
               names_prefix = "bill_",
               names_to = c("bill_measure" , "unit"),
               names_sep = "_")
```

```{r pivot-l-3c-check}
grade_code(
  correct = random_praise(),
  incorrect = random_encouragement()
)
```

```{r pivot-l-3c-hint-1}
Now that the "bill_" prefix is removed, there are only two columns produced and not three!
```

```{r pivot-l-3c-hint-2}
To be clear about the column content, name the first pivoted column "bill_measure".
```



## Challenge 4

### 4a
> Turn the penguins_long_simple dataset back to its original state

```{r pivot-w-4a, exercise = TRUE}
penguins_long_simple %>% 
  pivot_wider(_ = _,
              _ = _)
```

```{r pivot-w-4a-solution}
penguins_long_simple %>% 
  pivot_wider(names_from = name,
              values_from = value)
```

```{r pivot-w-4a-check}
grade_code(
  correct = random_praise(),
  incorrect = random_encouragement()
)
```

```{r pivot-w-4a-hint-1}
Make sure you spell all column names correctly
```


## Challenge 5

### 5a
> Turn the penguins_long dataset back to its original state

```{r pivot-w-5a, exercise = TRUE}
penguins_long %>% 
  pivot_wider(_ = _,
              _ = _,
              _ = _)
```

```{r pivot-w-5a-solution}
penguins_long %>% 
  pivot_wider(names_from = c("part", "measure", "unit"),
              names_sep = "_",
              values_from = value)
```

```{r pivot-w-5a-check}
grade_code(
  correct = random_praise(),
  incorrect = random_encouragement()
)
```

```{r pivot-w-5a-hint-1}
Make sure you spell all column names correctly
```


<!-- ## Quiz -->

<!-- ```{r quiz} -->
<!-- quiz( -->
<!--   question("What functions can you use to subset a data set by rows?", -->
<!--     answer("dplyr's `filter()`", correct = TRUE), -->
<!--     answer("dplyr's `select()`"), -->
<!--     answer("`subset()`", correct = TRUE),
    allow_retry = TRUE -->
<!--   ), -->
<!--   question("What functions can you use to subset a data set by columns", -->
<!--     answer("dplyr's `filter()`"), -->
<!--     answer("dplyr's `select()`", correct = TRUE), -->
<!--     answer("`subset()`", correct = TRUE),
    allow_retry = TRUE -->
<!--   ), -->
<!--   question("If you want to select all columns in data 'df' that contains the string 'something', you can do that by", -->
<!--     answer("`df[grepl('something', names(df))]`", correct = TRUE), -->
<!--     answer("`select(df, starts_with('something')`"), -->
<!--     answer("`df[,'something']`"), -->
<!--     answer("`select(df, contains('something')`", correct = TRUE),
    allow_retry = TRUE -->
<!--   ), -->
<!--     question("If you want to subset rows so that you only have those below 18 years of age, you can do that by", -->
<!--     answer("`df$age < 18`"), -->
<!--     answer("`filter(df, age < 18)`", correct = TRUE), -->
<!--     answer("`df[df$age < 18,]`", correct = TRUE), -->
<!--     answer("`filter(df, age <= 18)`"),
    allow_retry = TRUE -->
<!--   )
<!-- ) -->
<!-- ``` -->



