---
title: "Working with datasets in R and the Tidyverse"
subtitle: "Visualisation with ggplot2"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: true
date: "25th May 2021"
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      fig.retina = 3)
```

> **Questions**
>
> How can I create publication-quality graphics in R?
>
> **Objectives**
>
> To be able to use `ggplot2` to generate publication quality graphics.
>
> To understand the basic grammar of graphics, including the aesthetics and geometry layers, adding statistics, transforming scales, and colouring or panelling by groups.

Plotting the data is one of the best ways to quickly explore it and generate hypotheses about various relationships between variables.

There are several plotting systems in R, but today we will focus on `ggplot2` which implements **grammar of graphics** - a coherent system for describing components that constitute visual representation of data. For more information regarding principles and thinking behind `ggplot2` graphic system, please refer to [Layered grammar of graphics](http://vita.had.co.nz/papers/layered-grammar.pdf) by Hadley Wickham (@hadleywickham). 

The advantage of `ggplot2` is that it allows R users to create publication quality graphics with just a few lines of code. `ggplot2` has a large user base and is constantly developed and extended by the community.

# Getting started

`ggplot2` is a core member of `tidyverse` family of packages. Installing and loading the package under the same name will load all of the packages we will need for this workshop. Lets get started!

```{r}
# install.packages("tidyverse")
# install.packages("penguins")
library(tidyverse)
```

If above code produces an error "there is no package called ‘tidyverse’", uncomment (remove #) the line above and run `install.packages()`command before you load the library. You only need to install the package once, but you will have to reload it, using the `library()` command, every time you restart R. 

Today we will be working with the `penguins` dataset, which is the excerpt from the [penguins](https://allisonhorst.github.io/palmerpenguins/) data. Data were collected and made available by [Dr. Kristen Gorman](https://www.uaf.edu/cfos/people/faculty/detail/kristen-gorman.php) and the [Palmer Station, Antarctica LTER](https://pal.lternet.edu/), a member of the [Long Term Ecological Research Network](https://lternet.edu/). 

First let's load the data. 

```{r}
penguins <- palmerpenguins::penguins
```

Here, we grab the data called `penguins` from the palmerpenguins package, and assign (`<-`) it to an `object` in our R environment we call `penguins`.
Notice how you can see it in your environment pane in RStudio.
You can have a look at the content of the `penguins` data frame by simply typing `penguins` either in the R-chunk or in the console. Data frame is a rectangular collection of data, where variables are organized as columns and observations are listed as rows.

```{r}
penguins
```


The dataset contains the following fields:

- **species**:           penguin species
- **island**:            island of observation
- **bill_length_mm**:    bill length in millimetres
- **bill_depth_mm**:     bill depth in millimetres
- **flipper_length_mm**: flipper length in millimetres
- **body_mass_g**:       body mass in grams
- **sex**:               penguin sex
- **year**:              year of observation

More information about the package and the data is available in help. Just type `?penguins` in console, located in the bottom panel of your RStudio, or type `penguins` in the search field of the *Help* tab of the bottom-right RStudio panel. Whenever you are unsure about anything in R, it is a good idea to check out the help file using one of the two methods described above.

# Creating the first plot

> Here's a question that we would like to answer using `penguins` data: _Do penguins with high body mass also have long beaks?_ This might seem like a silly question, but it gets us exploring our data.

To plot `penguins`, run the following code in the R-chunk or in console. The following code will put `body_mass_g` on the x-axis and `bill_length_mm` on the y-axis:

```{r}
ggplot(data = penguins) + 
  geom_point(
    mapping = aes(x = body_mass_g,
                  y = bill_length_mm)
  )
```

Note that we split the function into several lines. 
In R, any function has a name and is followed by parentheses. Inside the parentheses we place any information the function needs to run.
Here, we are using two main functions, ggplot() and geom_point().
To save screen space, we have placed each function on its own line, and also split up arguments into several lines.
How this is done depends on you, there are no real rules for this.
We will use the tidyverse coding style throughout this course, to be consistent and also save space on the screen.
The plus sign indicates that the ggplot is not over yet and that the next line should be interpreted as additional layer to the preceding `ggplot()` function. In other words, when writing a `ggplot()` function spanning several lines, the `+` sign goes at the end of the line, not in the beginning.

The plot shows positive linear relationship between bill length and body mass. 

> Does this graph confirm or disprove your initial hypothesis about the relationship between these variables?

Note that in order to create a plot using `ggplot2` system, you should start your command with `ggplot()` function. It creates an empty coordinate system and initializes the dataset to be used in the graph (which is supplied as a first argument into the `ggplot()` function). In order to create graphical representation of the data, we can add one or more layers to our otherwise empty graph. Functions starting with the prefix `geom_` create a visual representation of data. In this case we added scattered points, using `geom_point()` function. There are many `geoms` in `ggplot2`, some of which we will learn in this lesson.

`geom_` functions create *mapping* of variables from the earlier defined dataset to certain aesthetic elements of the graph, such as axis, shapes or colours. The first argument of any `geom_` function expects the user to specify these mappings, wrapped in the `aes()` (short for *aesthetics*) function. In this case, we mapped `body_mass_g` and `bill_length_mm` variables from `penguins` dataset to x and y-axis, respectively (using `x` and `y` arguments of `aes()` function). 

Generally speaking, the template for visualizing data in `ggplot2` can be summarized as follows:


```{r, eval=FALSE}
`ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))`
```

In the remainder of this lesson we will learn how to extend and complete this template using different elements to produce various visualizations. First, we will look closer at the `<MAPPINGS>` component.

## **Challenge 1.** {.tabset}

You can run assignments in your own RStudio, or run the first challenge in the plotting tutorial by entering the following in the R console:
```r
learnr::run_tutorial("001-ggplot2", "swc.tidyverse")
```
(helpers, please paste this into the chat at the right time.)

### Assignment
>
> 1a: How does body mass change over time? What do you observe? Note that many points are plotted on top of each other. This is called "overplotting". 
>
> *Hint: the* `penguins` *dataset has a column called `year`, which should appear on the x-axis.*
>
> 1b: Try a different `geom_` function called `geom_jitter`. It will spread the points apart a little bit using random noise.
>
> 1c: See if you can visualize body mass by island. Which island tends to have higher body mass (notice the density of the points along the y-axis)? Lowest body mass? Which island has highest spread in body mass values? How about lowest spread? 

### Solution

```{r, results="hide"}
## 1a
ggplot(data = penguins) +
  geom_point(
    mapping = aes(x = year, 
                  y = bill_length_mm)
  )

# 1b
ggplot(data = penguins) +
  geom_jitter(
    mapping = aes(x = year, 
                  y = bill_length_mm)
  )

## 1c
ggplot(data = penguins) +
  geom_point(
    mapping = aes(x = island, 
                  y = bill_length_mm)
  )

```


# Mapping data pt.1

What if we want to combine graphs from the previous two challenges and show the relationship between three variables in the same graph? Turns out, we don't necessarily need to use third geometrical dimension, we can simply employ colour.

The following graph maps `island` variable from `penguins` dataset to the `colour` aesthetic of the plot. Let's take a look:

```{r}
ggplot(data = penguins) + 
  geom_jitter(
    mapping = aes(x = year, 
                  y = bill_length_mm, 
                  colour = island)
  )
```

## **Challenge 2.** {.tabset}
You can run assignments in your own RStudio, or run the second challenge in the plotting tutorial by entering the following in the R console:
```r
learnr::run_tutorial("001-ggplot2", "swc.tidyverse")
```
(helpers, please paste this into the chat at the right time.)

### Assignment
> 
> **2a**: What will happen if you switch the mappings of `island` and `year` in the previous example? Is the graph still useful? Why? Try mapping year to colour.
>
> **2b**: What if you map `colour` aesthetic to `species`? What has changed? How is `year` different from `species`? What is the limitation of the `colour` aesthetic, when used to visualize different types of data? 
> 
> **2c**: Can you add a little colour to our initial graph of body mass by bill length? colour the points by island.
>
> **2d**: How about using colour gradient to illustrate change over time?


### Solution

```{r, results="hide"}
## 2a
ggplot(data = penguins) + 
  geom_jitter(
    mapping = aes(x = bill_length_mm, 
                  y = year,
                  colour = year)
  )

# 2b
ggplot(data = penguins) + 
  geom_jitter(
    mapping = aes(x = island, 
                  y = bill_length_mm,
                  colour = species)
  )


## 2c
ggplot(data = penguins) + 
  geom_jitter(
    mapping = aes(x = body_mass_g, 
                  y = bill_length_mm,
                  colour = island)
  )

# 2d
ggplot(data = penguins) + 
  geom_jitter(
    mapping = aes(x = body_mass_g, 
                  y = bill_length_mm,
                  colour = year)
  )
```


# Mapping data pt.2

There are other aesthetics that can come handy. One of them is `size`. The idea is that we can vary the size of data points to illustrate another continuous variable, such as species bill depth. Lets look at four dimensions at once! 

```{r}
ggplot(data = penguins) + 
  geom_point(
    mapping = aes(x = body_mass_g, 
                  y = bill_length_mm, 
                  colour = island, 
                  size = bill_depth_mm)
  )
```

There's one more useful aesthetic property of the graph which is good for visualizing low-cardinality categorical variables (categorical variables with small number of unique values), called `shape`. The idea is that you can employ different shapes (other than circles) to plot the data.

## **Challenge 3.** {.tabset}
You can run assignments in your own RStudio, or run the third challenge in the plotting tutorial by entering the following in the R console:
```r
learnr::run_tutorial("001-ggplot2", "swc.tidyverse")
```
(helpers, please paste this into the chat at the right time.)

### Assignment
> 
> **3**: Blow your mind by visualizing five(!) dimensions in the same graph. Modify the previous example mapping year to colour and shape to island. 

### Solution

```{r, results="hide"}
ggplot(data = penguins) + 
  geom_point(
    mapping = aes(x = body_mass_g, 
                  y = bill_length_mm, 
                  colour = year, 
                  shape = island,
                  size = bill_depth_mm)
  )
```

# Setting values

Combining too many aesthetics in the same graph can make it quite busy. However, you can always remove certain aesthetic properties and use several graphs to highlight different aspects of data.

Until now, we explored different aesthetic properties of a graph mapped to certain variables. What if you want to recolour or use a certain shape to plot all data points? Well, that means that such colour or shape will no longer be *mapped* to any data, so you need to supply it to `geom_` function as a separate argument (outside of the `mapping`). 
This is called "setting" in the ggplot2-world. We "map" aesthetics to data columns, or we "set" single values outside aesthetics to apply to the entire geom or plot.
Here's our initial graph with all colours coloured in blue.

```{r}
ggplot(data = penguins) + 
  geom_point(
    mapping = aes(x = body_mass_g, 
                  y = bill_length_mm),
    colour = "blue"
  )
```

Once more, observe that the colour is now not mapped to any particular variable from the `penguins` dataset and applies equally to all data points, therefore it is outside the `mapping` argument and is not wrapped into `aes()` function. Note that set colours are supplied as characters (in quotes). 

## **Challenge 4.** {.tabset}

You can run assignments in your own RStudio, or run the forth challenge in the plotting tutorial by entering the following in the R console:
```r
learnr::run_tutorial("001-ggplot2", "swc.tidyverse")
```
(helpers, please paste this into the chat at the right time.)

### Assignment
> 
> **4a**: Try mapping `colour` aesthetic to `island` and then to `year`. What do you notice? What might be the reason for different treatment of these variables by `ggplot`?
> 
> **4b**: Change the transparency of the data points by year.
> 
> **4c**: Move the transparency outside the `aes()` and set it to `0.7`. What can be the benefit of each one of these methods?
> 
> **4d**: Add colour argument, with 'blue' in quotations, into the aes and see what happens. Did you expect that?

### Solution
```{r, results="hide"}
## 4a
ggplot(data = penguins) + 
  geom_point(
    mapping = aes(x = body_mass_g, 
                  y = bill_length_mm, 
                  colour = year)
  )

## 4b
ggplot(data = penguins) + 
  geom_point(
    mapping = aes(x = body_mass_g, 
                  y = bill_length_mm, 
                  alpha = year)
  )

## 4c
ggplot(data = penguins) + 
  geom_point(
    mapping = aes(x = body_mass_g, 
                  y = bill_length_mm),
    alpha = 0.7)

## 4d
ggplot(data = penguins) + 
  geom_point(
    mapping = aes(x = body_mass_g, 
                  y = bill_length_mm,
                  colour = "blue")
  )
```

When an argument is placed inside an aes and remains quoted, like "red" here, ggplot is interpreting as a variable named "blue" and not the colour blue!


# Labeling the chart

Lastly we will learn how to label and annotate the chart using `labs` and `annotate` functions. 

```{r}
ggplot(data = penguins) + 
  geom_point(
    aes(x = body_mass_g, 
        y = bill_length_mm, 
        colour = island)
  ) +
  facet_wrap(~year) + 
  labs(
    title="Body mass vs bill length over time",
    subtitle="In the past 3 years, body mass of penguins do not seem to increase.",
    caption="Originally published in doi:10.1371/journal.pone.0090081",
    x = "Bill length, (mm)",
    y = "Body mass (grams)",
    colour = "Island"
  )
```


# Geometrical objects

Next, we will consider different options for <GEOM_FUNCTION> component of our template `ggplot2` graph example. Using different `geom_` functions user can highlight different aspects of data. For example, we could connect individual data points belonging to the same species into a line and illustrate the development of body mass over time for each species separately using `geom_line()` function. 

Some `geom_` functions allow additional aesthetics, such as aesthetic `group` in the `geom_line()` function. This aesthetic may not have any meaning in other geoms, but here it allows us to draw multiple lines, one per species. To keep the lines organized, we will colour them by island. 

```{r}
ggplot(data = penguins) + 
  geom_line(
    mapping = aes(x = year, 
                  y = bill_length_mm, 
                  group = species, 
                  colour = species)
  )
```

That looks crazy! Line plots are not ideal for this data, as there are multiple lines per speces and year, so the lines do not line up nicely.

Another useful geom function is `geom_boxplot()`. It adds a layer with the "box and whiskers" plot illustrating the distribution of values within categories. The following chart breaks down body mass by island, where the box represents first and third quartile (the 25th and 75th percentiles), the middle bar signifies the median value and the whiskers extent to cover 95% confidence interval. Outliers (outside of the 95% confidence interval range) are shown separately.

```{r}
ggplot(data = penguins) + 
  geom_boxplot(
    mapping = aes(x = island, 
                  y = bill_length_mm)
  )
```

Layers can be added on top of each other. In the following graph we will place the boxplots over jittered points to see the distribution of outliers more clearly. We can map two aesthetic properties to the same variable. Here we will also use different colour for each island.

```{r}
ggplot(data = penguins) + 
  geom_jitter(
    mapping = aes(x = island, 
                  y = bill_length_mm, 
                  colour = island)
  ) +
  geom_boxplot(
    mapping = aes(x = island,
                  y = bill_length_mm, 
                  colour = island)
  )

```

Now, this was slightly inefficient due to duplication of code - we had to specify the same mappings for two layers. To avoid it, you can move common arguments of `geom_` functions to the main `ggplot()` function. In this case every layer will "inherit" the same arguments, specified in the "parent" function.

```{r}
ggplot(data = penguins,
       mapping = aes(x = island, 
                     y = bill_length_mm,
                     colour = island)
) + 
  geom_jitter() +
  geom_boxplot()

```

You can still add layer-specific mappings or other arguments by specifying them within individual geoms. We would recommend building each layer separately and then moving common arguments up to the "parent" function.

We can use linear models to highlight differences in dependency between bill length and body mass by island. Notice that we added a separate argument to the `geom_smooth()` function to specify the type of model we want `ggplot2` to built using the data (linear model). The `geom_smooth()` function has also helpfully provided confidence intervals, indicating "goodness of fit" for each model (shaded gray area). For more information on statistical models, please refer to help (by typing `?geom_smooth`)

```{r}
ggplot(data = penguins, 
       mapping = aes(x = body_mass_g, 
                     y = bill_length_mm,
                     colour = island)
) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm")
```

Notice, that we also used a new visual property called `alpha` to increase transparency of the data points and make trend lines stand out. `alpha` property can also be used as a mapping aesthetic, i.e. transparency can be made to vary depending on the value of certain variable.

## **Challenge 5.** {.tabset}
You can run assignments in your own RStudio, or run the fifth challenge in the plotting tutorial by entering the following in the R console:
```r
learnr::run_tutorial("001-ggplot2", "swc.tidyverse")
```
(helpers, please paste this into the chat at the right time.)

### Assignment
> 
> Modify the graph to force R to create single regression line for all data points. Keep the points coloured by island.
>

### Solution

In the graph above, each geom inherited all three mappings: x, y and colour. If we want only single linear model to be built, we would need to limit the effect of `colour` aesthetic to only `geom_point()` function, by moving it from the "parent" function to the layer where we want it to apply. Note, though, that because we want the `colour` to be still mapped to the `island` variable, it needs to be wrapped into `aes()` function and supplied to `mapping` argument.

```{r, results="hide"}
ggplot(data = penguins, 
       mapping = aes(x = body_mass_g, 
                     y = bill_length_mm)
) +
  geom_point(mapping = aes(colour = island),
             alpha = 0.5) +
  geom_smooth(method = "lm")

# Alternative solution
ggplot(data = penguins, 
       mapping = aes(x = body_mass_g, 
                     y = bill_length_mm, 
                     colour = island)
) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", 
              colour = "black")
```

# Correcting the scale

Some times, we have data that need to be plotted on another scale than the default native scale of the data.
For instance, if you have data that have some type of exponential or hyperbolic distribution.
There are a couple of ways to do this in ggplot2. 
The first thing we can try is transforming the data with a function, like the `log()` for a logarithmic transformation.

```{r}
ggplot(data = penguins,
       mapping = aes(x = log(body_mass_g), 
                     y = bill_length_mm, 
                     colour = island)
) +
  geom_point() +
  geom_smooth(method = "lm")
```

As you can observe the x-axis label of our graph says `body_mass_g`, which indicates that we are not really plotting the original data, but rather the output of `log()` function. The same effect (with slightly more aesthetically pleasing x-axis label) can be achieved by specifying the x-axis scale transformation as a separate layer. Instead of transforming the values, we will transform the scale of x-axis.

```{r}
ggplot(data = penguins,
       mapping = aes(x = body_mass_g, 
                     y = bill_length_mm, 
                     colour = island)
) +
  geom_point() +
  geom_smooth(method = "lm") +
  scale_x_log10()
```

Now the x-axis is measured in log10 units and the data, plotted on log10 scale looks more linear. Certain scale and coordinate functions may result in similar visual effects on the chart, but the way they interact with other aesthetic elements may be quite different. Check out the online `ggplot2` [documentation](http://docs.ggplot2.org/) for more details and examples of using scale and coordinate transformations.

## **Challenge 6.** {.tabset}
You can run assignments in your own RStudio, or run the sixth challenge in the plotting tutorial by entering the following in the R console:
```r
learnr::run_tutorial("001-ggplot2", "swc.tidyverse")
```
(helpers, please paste this into the chat at the right time.)

### Assignment
> 
> **6a**: Make a boxplot of body mass by year. When was interquartile range of body mass the smallest? 
>
> **6b**: Make a histogram of `body_mass_g`. What is the shape of the distribution? Try setting bin to 50. Why is the bin parameter important for interpretation of the histogram? 
>
> **6c**:  Build a density function. How would you compare density functions of different islands?
> 
> **6d**: Based on graph produced using `geom_density2d()` function of log bill length vs body mass, how many clusters of data points can you identify? What if you look at it by island?

### Solution
```{r, results="hide"}
## 6a
ggplot(penguins) +
  geom_boxplot(
    mapping = aes(y = body_mass_g, 
                  x = year, 
                  group = year)
  )

## 6b
ggplot(penguins) +
  geom_histogram(
    mapping = aes(x = body_mass_g),
    bins = 50
  )

## 6c
ggplot(penguins) +
  geom_density(
    mapping = aes(x = body_mass_g, 
                  colour = island)
  ) 

## 6d
ggplot(penguins) +
  geom_density2d(
    mapping = aes(x = body_mass_g,
                  y = bill_length_mm,
                  colour = island)
  )
```

# Faceting

Multi-layered graphs employing several aesthetics can look crowded. In order to avoid it, one can split the data into different graphs using panels of similar graphs. In `ggplot2` this method is called "faceting". Lets facet the graph above by island and show the data points and the trend for each island in a separate chart.

```{r}
ggplot(data = penguins, 
       mapping = aes(x = body_mass_g, 
                     y = bill_length_mm)
) +
  geom_point() +
  geom_smooth() +
  facet_wrap(~ island)
```

The `facet_wrap()` layer takes a “formula” as its argument, denoted by the tilde (~). This tells R to draw a panel for each unique value in the `island` column of the `penguins` dataset. Faceting is useful when number of panels is limited. Notice that here R places panels from left to right, "wrapping" those panels that do not fit in one row onto the new line. Learn about advanced faceting, including faceting over several variables using help on `?facet_grid()`.

Reiterating our previously proposed `ggplot2` template and adding what we learned until, now we can state:

```{r, eval=FALSE}
`ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>)) + 
  <FACET_FUNCTION>`
```

## **Challenge 7.** {.tabset}
You can run assignments in your own RStudio, or run the seventh challenge in the plotting tutorial by entering the following in the R console:
```r
learnr::run_tutorial("001-ggplot2", "swc.tidyverse")
```
(helpers, please paste this into the chat at the right time.)

### Assignment
> 
> **7a**: Try faceting by year, keeping the linear smoother. Is there any change in slope of the linear trend over the years?
>
> **7b**: What if you look at linear models per island?

### Solution

```{r, results="hide"}
## 7a
ggplot(data = penguins, 
       mapping = aes(x = body_mass_g, 
                     y = bill_length_mm)
) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~ year)

## 7b
ggplot(data = penguins, 
       mapping = aes(x = body_mass_g, 
                     y = bill_length_mm)
) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap( ~ island)
```

# Changing color and fill

Often, people would want to change the colours in the plots or the general theme to something more suitable to their needs. Journals often have specific requirements to how plots should look, or you need to use specific colours across different types of visualisations.

Changing the colours are done through the `scale_fill` and `scale_colour` functions.

```{r}
ggplot(penguins)+
  geom_density(
    aes(x = body_mass_g, 
        colour = island)
  ) +
  scale_colour_brewer(palette = "Dark2")
```

This plot has used colour on a density, which will colour the edges of the density plot. 
This is not very clear, however, and colours are not easy to see.
We can try switching the colour to "fill" mapping instead, and also change the scale to fill, which should flood the density area.

Can you notice something other that is different with this code?
We have taken away the `data = ` and `mapping = ` calls! ggplot is clever. In the `ggplot()` call, as long as we provide the data to use as the first input, it know this is data. And all the geoms auto-detect the `aes()`, so you dont _need_ to put `mapping =` there!

```{r}
ggplot(penguins)+
  geom_density(
    aes(x = body_mass_g, 
        fill = island), 
    alpha = .5
  ) +
  scale_fill_brewer(palette = "Dark2")
```
Here, we are using the [brewer palettes](http://applied-r.com/rcolorbrewer-palettes/#the-palettes) to colour. These have the very convenient `scale_[]_brewer()` functions we can use. 
We can also use our own palettes if we want!


```{r}
ggplot(penguins)+
  geom_density(
    aes(x = body_mass_g, 
        fill = island),
    alpha = .5
  ) +
  scale_fill_manual(values = c("firebrick", "dodgerblue", "forestgreen"))
```
Its often smart to use palettes that are curated though, as they often include colour scales that give good distinctions between colours. Another popular variant like brewer is the viridis colours.

```{r}
ggplot(penguins)+
  geom_density(
    aes(x = body_mass_g, 
        fill = island), 
    alpha = .5
  ) +
  scale_fill_viridis_d()
```

## **Challenge 8.** {.tabset}
You can run assignments in your own RStudio, or run the eigth challenge in the plotting tutorial by entering the following in the R console:
```r
learnr::run_tutorial("001-ggplot2", "swc.tidyverse")
```
(helpers, please paste this into the chat at the right time.)

### Assignment
> 
> **8a**: Make a boxplot of body mass by year. What happens if you add `factor()` around year? What do you need to change in the scale_fill function to make it work? 
>
> **8b**: Make a histogram of `body_mass_g`? What is the shape of the distribution? Why is bin parameter important for interpretation of the histogram? 
>
> **8c**: Build a density plot How would you compare density functions of different islands?
> 

### Solution
```{r, results="hide"}
## 8a
ggplot(penguins) +
  geom_boxplot(
    mapping = aes(y = body_mass_g, 
                  x = factor(year),
                  fill = factor(year))
  ) +
  scale_fill_viridis_d()

## 8b
ggplot(penguins) +
  geom_point(
    mapping = aes(x = body_mass_g,
                  y = bill_length_mm, 
                  colour = body_mass_g)
  ) +
  scale_colour_viridis_c()

## 8c
ggplot(penguins)+
  geom_density2d(
    aes(x = body_mass_g, 
        y = bill_length_mm, 
        colour = island)
  ) +
  scale_colour_brewer(palette = "Dark2")
```

# Themes

Now that we can change the colours, we might want to change the general plot look. Not everyone likes the grey grid background, the grid ticks etc. 

In ggplot2, we change this static part of the plot through the `theme` functions. There are several built in versions that have common layouts people might use. For instance, `theme_classic()` is often used when preparing for journal submissions for journals that have very specific requirements for plot composition.

```{r}
ggplot(penguins) +
  geom_boxplot(
    mapping = aes(y = body_mass_g, 
                  x = factor(year),
                  fill = factor(year))
  ) +
  scale_fill_viridis_d() +
  theme_classic()
```

In addition to the built in themes, you can control _everything_ in the plot layout through the simple `theme` call. While the function name is simple, altering theme elements can be quite difficult, as its quite advanced. But there are some specific arguments people often like to use, like repositioning where the legend appears.

```{r}
ggplot(penguins) +
  geom_boxplot(
    mapping = aes(y = body_mass_g, 
                  x = factor(year),
                  fill = factor(year))
  ) +
  scale_fill_viridis_d() +
  theme(
    legend.position = "bottom"
  )
```

## **Challenge 9.** {.tabset}
You can run assignments in your own RStudio, or run the ninth challenge in the plotting tutorial by entering the following in the R console:
```r
learnr::run_tutorial("001-ggplot2", "swc.tidyverse")
```
(helpers, please paste this into the chat at the right time.)

### Assignment
> 
> **9a**: Create a plot and alter the theme. Try the dark theme, for instance!
>
> **9b**: Edit the theme and make the plot as ugly as you can! Use both the theme and scales for the colours to find the most horrible combinations! 
> 

### Solution
```{r, results="hide"}
## 9a
ggplot(penguins) +
  geom_boxplot(
    mapping = aes(y = body_mass_g, 
                  x = factor(year),
                  fill = factor(year))
  ) +
  theme_dark()

## 9b
ggplot(penguins)+
  geom_density2d(
    aes(x = body_mass_g, 
        y = bill_length_mm, 
        colour = island)
  ) +
  scale_colour_brewer(palette = "Dark2")
```


# Wrap-up

We conclude this lesson by reiterating our `ggplot2` data visualization template. 

```{r, eval=FALSE}
`ggplot(data = <DATA>,
        mapping = aes(<GLOBAL_MAPPINGS>)) + 
  <GEOM_FUNCTION>(
    mapping = aes(<GEOM_MAPPINGS>)
  ) +
  <SCALE_FUNCTION> +
  <COORDINATE_FUNCTION> +
  <FACET_FUNCTION> + 
  <LABS>`
```

We learned about seven parameters of ggplot functions. However, it is very rare that all six of them need to specified in a given graphic or chart. Most of the time ggplot offers useful defaults for everything other than data, geoms and mappings.

